Automating Free Logic in HOL, with an Experimental Application in Category Theory

### 2018-10-14 FreeLogic/2018-JAR/article.tex
corrections

  A shallow semantical embedding of free logic in classical
  higher-order logic is presented, which enables the off-the-shelf
  application of higher-order interactive and automated theorem
  provers for the formalisation and
  verification of free logic theories.  
  Subsequently, this approach is applied to a selected domain of
  mathematics:
 starting from a generalization of the standard axioms
  for a monoid we present a stepwise development of various, mutually
  equivalent foundational axiom systems for category theory. As a
  side-effect of this work some (minor) issue in a prominent category
  theory textbook has been revealed.

@@

It is obvious how this can be used to model undefideness and
partiality: problematic terms, e.g. division by zero or improper
definite descriptions, still denote, but they refer to undefined
objects, that is, objects $d$ in $\D\setminus \ex$ lying outside of
the scope of quantification.
%In this framework partial and total functions are modelled as follows: 
Moreover, a function $f$ is \emph{total} if and only if for all $x$ we
have $\ex x \imp \ex(f x)$.\footnote{The predication $\ex x$ represents that $x$ is a
  member of $E$.}
For \emph{partial} functions $f$ we may have some $x$ such that $\ex x$ but not
$\ex(f x)$. A function $f$ is \emph{strict}  if  and only if for all $x$
 we have $\ex(f x) \imp \ex x$.

The particular notion of free logic as exploited in the remainder of
this article has been proposed by Scott \cite{Scott67}.  A graphical
illustration of this notion of free logic is presented in
Fig.~\ref{fig1}. It
employs a distinguished undefined object $\star$.\footnote{The $\star$
  symbol is not to be confused with any other symbol in Isabelle/HOL.}
% Free logic (and inclusive logic) has been proposed as an alternative
% to remedy these shortcomings. It distinguishes between a raw domain of
% possibly non-existing objects ‹❙D› and a particular subdomain ‹❙E› of ‹❙D›,
% containing only the ``existing'' entities. Free variables range over ‹❙D›
% and quantified variables only over ‹❙E›. Each term denotes in ‹❙D› but not
% necessarily in ‹❙E›. The particular notion of free logic as exploited below has been
% introduced by Scott @{cite "scott67:_exist"}. A graphical 
% illustration of this notion of free logic is presented in Fig.~\ref{fig1}.

@@

We formally introduce the syntax and semantics of free logic as
assumed in the remainder of this article. We refer to this logic with $\FFOL$.

@@

\subsection{Classical Higher-Order Logic}
Simple type theory, also referred to as classical higher-order logic
(HOL) \cite{AndrewsSEP}, is an expressive logic formalism which is based on the simply typed
$\lambda$-calculus~\cite{barendregt13book}. 
  %\footnote{
    %Although \sys~actually supports a polymorphic typed $\lambda$-calculus
    %we will
    %stick here to the simply typed $\lambda$-calculus for a better understanding.}
  HOL has its origin in the work by
  Church~\cite{Church40}.

For a detailed discussion of the syntax, semantics and automation
of  HOL we refer to the literature (see e.g.
\cite{AndrewsSEP,B5,J6} and the
references therein). Below we introduce a variant of \HOL with primitive equality
and definite description.

@@

The different properties of $\FFOL$ could now be experimentally explored with
automated reasoning tools for $\HOL$. We have conducted such
experiments in 
an earlier paper \cite{ICMS}. These experiments confirm the
illustrating examples discussed in Scott's paper \cite{Scott67}.

\section{Exploring Axioms Systems for Category Theory}
In an experimental theory exploration study, utilizing the
free logic reasoning framework from above, 
we have shown how Scott's
\cite{Scott79} axiom system for category theory can be derived from a
notion of partial monoids. These axioms systems are presented in Table \ref{axioms-sets-1}.

@@

  In our experiments we have also made use of the Isabelle/HOL's
  \textsc{smt} method, which ``translates the conjecture and any
  user-supplied facts to the SMT solvers’ many-sorted first-order
  logic, invokes a solver, and (depending on the solver) either trusts
  the result or attempts to reconstruct the proof in Isabelle.''
  (\cite[p.~5]{Sledgehammer}).\footnote{Technical remark: We have
    selected CVC4 in our experiments as the default SMT solver, since
    we did run into errors when working with Z3. These errors can
    easily be reconstructed in the provided source files when
    switching back to Z3 as default.}  For quite some time the use of the \textsc{smt}
  method has been controversially discussed in the Isabelle/HOL
  community. However, since about a year, the \textsc{smt} method is
  now even eligible in submissions to the Archive of Formal
  Proofs. For the work presented here this community internal discussion is 
  of little relevance, since we could replace all uses of the\textsc{smt} method
  by alternative proofs (which we avoid in order to keep
  our formalisation concise).

@@

\begin{table} \centering \normalsize
\begin{tabular}{ll}
\hline
\\
\multicolumn{2}{c}{Axioms Set FS-I:  Freyd and Scedrov in original notation (with issues)} \\
%\hline
\\
  $A1$ &  $\ex(x \circ y) \longleftrightarrow (x\Box \cong \Box y)$ \\ 
  $A2a$ & $((\Box x)\Box) \cong \Box x$ \\ 
  $A2b$ & $\Box (x\Box) \cong \Box x$ \\ 
  $A3a$ & $(\Box x) \circ x \cong x$ \\ 
  $A3b$ & $x \circ (x\Box) \cong x$ \\ 
  $A4a$ & $\Box(x \circ y) \cong \Box(x \circ (\Box y))$ \\ 
  $A4b$ & $(x \circ y)\Box \cong ((x\Box) \circ y)\Box$ \\ 
  $A5$   & $x \circ (y \circ z) \cong  (x \circ y) \circ z$ \\
\\
\hline
\\
\multicolumn{2}{c}{Axioms Set FS-II: Freyd and Scedrov in our notation (with issues)} \\
%\hline
\\
  $A1$  & $\ex (x \comp y) \longleftrightarrow \dom x \cong \cod y$ \\
  $A2a$ & $\cod(\dom x) \cong \dom x$ \\  
  $A2b$ & $\dom(\cod y) \cong \cod y$ \\  
  $A3a$  & $x \comp (\dom x) \cong x$ \\ 
  $A3b$ & $(\cod y) \comp y \cong y$ \\
  $A4a$ & $\dom(x \comp y) \cong \dom((\dom x) \comp y)$ \\ 
  $A4b$ & $\cod(x \comp y) \cong \cod(x \comp (\cod y))$ \\ 
  $A5$   & $x \comp (y \comp z) \cong  (x \comp y) \comp z$   \\
\\
\hline
\\
\multicolumn{2}{c}{Axioms Set VI: Freyd and Scedrov in our notation
  and corrected} \\
%\hline
\\
  $A1'$  & $\ex (x \comp y) \longleftrightarrow \dom x \simeq\cod y$ \\
  $A2a$ & $\cod(\dom x) \cong \dom x$ \\  
  $A2b$ & $\dom(\cod y) \cong \cod y$ \\  
  $A3a$  & $x \comp (\dom x) \cong x$ \\ 
  $A3b$ & $(\cod y) \comp y \cong y$ \\
  $A4a$ & $\dom(x \comp y) \cong \dom((\dom x) \comp y)$ \\ 
  $A4b$ & $\cod(x \comp y) \cong \cod(x \comp (\cod y))$ \\ 
  $A5$   & $x \comp (y \comp z) \cong  (x \comp y) \comp z$   \\
\\
\hline
\end{tabular}
\caption{The axioms set of Freyd and Scedrov in their and our
  notation, together with a proposed correction.\label{axioms-sets-2}}
\end{table}

@@

Further work includes the extension of our work towards an embedding of
free higher-order logic, the continuation of our formalization studies in
category theory and the application of free logic to various other
mathematical domains, including, for example,  projective geometry.
Regarding extensions towards free  higher-order logic some first steps
have already been taken \cite{W56,makarenko16:_autom_logik_logik_stufe}.

>>>>

  A shallow semantical embedding of free logic in classical
  higher-order logic is presented, which enables the off-the-shelf
  application of higher-order interactive and automated theorem
  provers for the formalisation and
  verification of free logic theories.  
  Subsequently, this approach is applied to a selected domain of
  mathematics:
 starting from a generalization of the standard axioms
  for a monoid we present a stepwise development of various, mutually
  equivalent foundational axiom systems for category theory. As a
  side-effect of this work some (minor) issues in a prominent category
  theory textbook have been revealed.

@@

It is obvious how this can be used to model undefideness and
partiality: problematic terms, e.g. division by zero or improper
definite descriptions, still denote, but they refer to undefined
objects, that is, objects $d$ in $\D\setminus \ex$ lying outside of
the scope of quantification.
%In this framework partial and total functions are modelled as follows: 
Moreover, a function $f$ is \emph{total} if and only if for all $x$ we
have $\ex x \imp \ex(f x)$.\footnote{The predication $\ex x$ represents that $x$ is a
  member of $E$.}
For \emph{partial} functions $f$ we may have some $x$ such that $\ex x$ but not
$\ex(f x)$. A function $f$ is called \emph{strict}  if  and only if
for all $x\in \D$
 we have $\ex(f x) \imp \ex x$.

The particular version of free logic as exploited in the remainder of
this article was proposed by Scott \cite{Scott67}.  A graphical
illustration of this notion of free logic is presented in
Fig.~\ref{fig1}. It
employs a distinguished undefined object $\star$.\footnote{The $\star$
  symbol is not to be confused with any other symbol in Isabelle/HOL.}
% Free logic (and inclusive logic) has been proposed as an alternative
% to remedy these shortcomings. It distinguishes between a raw domain of
% possibly non-existing objects ‹❙D› and a particular subdomain ‹❙E› of ‹❙D›,
% containing only the ``existing'' entities. Free variables range over ‹❙D›
% and quantified variables only over ‹❙E›. Each term denotes in ‹❙D› but not
% necessarily in ‹❙E›. The particular notion of free logic as exploited below has been
% introduced by Scott @{cite "scott67:_exist"}. A graphical 
% illustration of this notion of free logic is presented in Fig.~\ref{fig1}.

@@

We next  formally introduce the syntax and semantics of free logic as
to be used in the remainder of this article. We refer to this logic as
$\FFOL$.

@@

\subsection{Classical Higher-Order Logic}
Simple type theory, also referred to as classical higher-order logic
(HOL) \cite{AndrewsSEP}, is an expressive logic formalism which is based on the simply typed
$\lambda$-calculus~\cite{barendregt13book}. 
  %\footnote{
    %Although \sys~actually supports a polymorphic typed $\lambda$-calculus
    %we will
    %stick here to the simply typed $\lambda$-calculus for a better understanding.}
  HOL has its origin in the work of
  Church~\cite{Church40}.

For a detailed discussion of the syntax, semantics and automation
of  HOL we refer to the literature (see e.g.
\cite{AndrewsSEP,B5,J6} and the
references therein). Below we introduce a variant of \HOL with primitive equality
and definite descriptions.

@@

The different properties of $\FFOL$ could now be experimentally explored with
automated reasoning tools for $\HOL$. We have conducted such
experiments in 
an earlier paper \cite{ICMS}. These experiments confirm the
illustrative examples discussed in Scott's paper \cite{Scott67}.

\section{Exploring Axioms Systems for Category Theory}
In an experimental theory-exploration study, utilizing the
free logic reasoning framework from above, 
we have shown how Scott's
\cite{Scott79} axiom system for category theory can be derived from a
notion of partial monoids. These axioms systems are presented in Table \ref{axioms-sets-1}.

@@

  In our experiments we have also made use of the Isabelle/HOL's
  \textsc{smt} method, which ``translates the conjecture and any
  user-supplied facts to the SMT solvers’ many-sorted first-order
  logic, invokes a solver, and (depending on the solver) either trusts
  the result or attempts to reconstruct the proof in Isabelle.''
  (\cite[p.~5]{Sledgehammer}).\footnote{Technical remark: We have
    selected CVC4 in our experiments as the default SMT solver, since
    we did run into errors when working with Z3. These errors can
    easily be reconstructed in the provided source files when
    switching back to Z3 as default.}  For quite some time the use of the \textsc{smt}
  method has been controversially discussed in the Isabelle/HOL
  community. However, since about a year, the \textsc{smt} method is
  now even eligible in submissions to the Archive of Formal
  Proofs. For the work presented here this community internal discussion is 
  of little relevance, since we could replace all uses of the \textsc{smt} method
  by alternative proofs (which we avoid in order to keep
  our formalisation concise).

@@

\begin{table} \centering \normalsize
\begin{tabular}{ll}
\hline
\\
\multicolumn{2}{c}{Axioms Set FS-I:  Freyd and Scedrov in original notation (with issues)} \\
%\hline
\\
  $A1$ &  $\ex(x \circ y) \longleftrightarrow (x\Box \cong \Box y)$ \\ 
  $A2a$ & $((\Box x)\Box) \cong \Box x$ \\ 
  $A2b$ & $\Box (x\Box) \cong \Box x$ \\ 
  $A3a$ & $(\Box x) \circ x \cong x$ \\ 
  $A3b$ & $x \circ (x\Box) \cong x$ \\ 
  $A4a$ & $\Box(x \circ y) \cong \Box(x \circ (\Box y))$ \\ 
  $A4b$ & $(x \circ y)\Box \cong ((x\Box) \circ y)\Box$ \\ 
  $A5$   & $x \circ (y \circ z) \cong  (x \circ y) \circ z$ \\
\\
\hline
\\
\multicolumn{2}{c}{Axioms Set FS-II: Freyd and Scedrov in our notation (with issues)} \\
%\hline
\\
  $A1$  & $\ex (x \comp y) \longleftrightarrow \dom x \cong \cod y$ \\
  $A2a$ & $\cod(\dom x) \cong \dom x$ \\  
  $A2b$ & $\dom(\cod y) \cong \cod y$ \\  
  $A3a$  & $x \comp (\dom x) \cong x$ \\ 
  $A3b$ & $(\cod y) \comp y \cong y$ \\
  $A4a$ & $\dom(x \comp y) \cong \dom((\dom x) \comp y)$ \\ 
  $A4b$ & $\cod(x \comp y) \cong \cod(x \comp (\cod y))$ \\ 
  $A5$   & $x \comp (y \comp z) \cong  (x \comp y) \comp z$   \\
\\
\hline
\\
\multicolumn{2}{c}{Axioms Set VI: Freyd and Scedrov in our notation
(corrected)} \\
%\hline
\\
  $A1'$  & $\ex (x \comp y) \longleftrightarrow \dom x \simeq\cod y$ \\
  $A2a$ & $\cod(\dom x) \cong \dom x$ \\  
  $A2b$ & $\dom(\cod y) \cong \cod y$ \\  
  $A3a$  & $x \comp (\dom x) \cong x$ \\ 
  $A3b$ & $(\cod y) \comp y \cong y$ \\
  $A4a$ & $\dom(x \comp y) \cong \dom((\dom x) \comp y)$ \\ 
  $A4b$ & $\cod(x \comp y) \cong \cod(x \comp (\cod y))$ \\ 
  $A5$   & $x \comp (y \comp z) \cong  (x \comp y) \comp z$   \\
\\
\hline
\end{tabular}
\caption{The axioms set of Freyd and Scedrov in their and our
  notation, together with a proposed correction.\label{axioms-sets-2}}
\end{table}

@@

Further work includes the extension of our work towards an embedding
of free higher-order logic, the continuation of our formalization
studies in category theory (especially extensions of the theory
involving functors) and the application of free logic to
various other mathematical domains, including, for example, projective
geometry.  Regarding extensions towards free higher-order logic some
first steps have already been taken
\cite{W56,makarenko16:_autom_logik_logik_stufe}, and a recent
continuation of our formalisation studies \cite{J39} now also includes 
an early axiom system for category theory by Saunders MacLane
\cite{maclane48:_group}.
